Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-28 paper/paper.qmd
update paper with prose error

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022. Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022. Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period. Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates. These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities. By utilizing recent, high-quality data and rigorous modeling, this study provides critical insights into the intersection of reproductive rights and infant health outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Instead of examining the impact of abortion laws over a broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database, captures infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a nuanced examination
of the effects of restrictive abortion laws on public health.

@@

This study is important as we highlight the public health implications
of abortion bans, particularly their role in increasing infant mortality
rates. By focusing on the 2022 Dobbs decision, we provide timely
evidence on how restrictive abortion policies disproportionately affect
vulnerable populations, emphasizing the need for data-driven policies to
mitigate these adverse outcomes.

@@

The process of translating real-world events into our dataset requires a
systematic approach to measurement and data gathering. In this research,
we analyze infant mortality rates in the United States to understand the
impact of abortion policies and demographic factors. The CDC’s Linked
Birth/Infant Death Records database provides comprehensive data,
including maternal and infant characteristics, to support this study.
Variables such as maternal age, race, education, birth weight,
gestational age, and prenatal care are included to ensure a detailed
examination of factors influencing infant health outcomes.

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
utilizing comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing the public
health strategies and legislative decisions.

@@

```{r}
#| label: fig-deathrate
#| fig-cap: The distribution of infant death rate (per 1,000 live births), with a peak observed between 5 and 10 deaths per 1,000 infants. The distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.
#| echo: false

@@

@@

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) +
  theme(legend.position = "bottom")
```

@fig-map illustrates the legality of abortion across U.S. states. Based
on @reproductiveright, as of November 2024, abortion has been illegal in
Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
Virginia, and Wisconsin following the June 2022 decision.

```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
#| message: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() +
  theme(legend.position = "bottom")
```

@@

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

The Year of Death variable represents the calendar year in which the
infant death occurred, specifically focusing on the years 2021 and 2022
in this study. This variable is essential for analyzing temporal trends
and identifying patterns or shifts in infant mortality rates over this
critical period. By controlling the year of death, we can closely
examine the impact of significant policy changes, such as the
overturning of Roe v. Wade in 2022, as well as other temporal factors on
infant health outcomes.

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
#| 
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

```{r}
#| label: tbl-motherrace
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age groups of Mothers. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

```{r}
#| label: fig-choromap
#| fig-cap: Infant Death Rates Across U.S. States (per 1,000 Live Births), December 2022. The infant Mortality Rate is much higher in Southeastern. Alabama and Mississippi show the highest rates, while California and New York exhibit lower rates.
#| echo: false

@@

By isolating the impact of the abortion injunction through this
interaction term `after_injunction`×`abortion_illegal`, our model
identifies changes in infant death rates that can be attributed to the
policy intervention. Additional details and diagnostics supporting the
model’s implementation is provided in [Appendix- @sec-model-details].

@@

For our analysis, we use separate Bayesian regression models to estimate
the infant death rates in states where abortion injunctions were
implemented and in states where abortion remains legal. We use the
interaction term between after_injunction and abortion_illegal to
measure the causal effect of the abortion injunction. Both models share
the same predictors to ensure consistency and comparability in capturing
the determinants of infant mortality.

For one of our models, we include control variables (`age_of_mother`,
`mothers_single_race`, and `year_of_death`) to adjust for confounders
that could influence infant death rates, addressing the assumption of
compositional differences in the DID approach and ensuring unbiased
estimates of the intervention’s effect. In the other model, we exclude
these controls focus on the overall intervention impact, and avoid
overfitting, and prioritize simplicity when variability is adequately
accounted for by the design or random effects. Thus, our decision to fit
two models to compare the outcomes: one adjusts for confounders to
refine the estimates, while the other focuses on the direct effect of
the intervention.

Our model includes `state` as a random effect to account for regional
variations in infant mortality that could introduce bias. We use a
random effect rather than a fixed effect to capture variability across
states without estimating specific effects for each one. This approach
models unobserved heterogeneity, such as differences in healthcare
systems and socioeconomic conditions, using a single variance term,
which avoids overfitting and ensures efficiency and generalizability. By
adjusting for unmeasured state-specific factors, the model provides
robust and reliable findings while balancing parsimony and
interpretability, making it well-suited for our analysis.

We choose the Bayesian framework since it enhances the DID approach by
incorporating prior information and quantifying uncertainty in a
probabilistic manner. In our Bayesian models, we use weakly informative
priors from the `rstanarm` package, including a Normal prior ($\mu=0$
and $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1)
for the auxiliary parameter, with `autoscale=TRUE` to dynamically adjust
priors based on the data scale. This enhances convergence and ensures priors
are calibrated to predictor ranges without dominating the data. The
model assumes plausible parameter centering, random effects capture
unobserved heterogeneity, and the DID framework relies on parallel
trends between groups. However, limitations include potential bias from
misspecified priors, computational challenges for hierarchical models,
and sensitivity to data sparsity or imbalances. This model may not be
suitable if group-level effects are negligible, parallel trends are
violated, or data is insufficient, but when applied appropriately, it
provides robust and interpretable causal estimates.

@@

@@

@@

The analysis reveals distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) reveals that abortion bans during the post-injunction
period results in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a critical role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

The findings highlight the significant public health consequences of
abortion bans in the United States. The observed increase of 0.285
infant deaths per 1,000 live births in states enforcing abortion bans
after the Dobbs decision underscores the detrimental impact of
restrictive abortion policies. These increases often reflect systemic
issues such as inadequate prenatal care, gaps in healthcare access, and
underlying socioeconomic inequities. States with pre-existing healthcare
challenges may experience amplified effects, suggesting an urgent need
for comprehensive policy interventions to address these disparities.
Tackling the root causes of these health disparities—such as improving
prenatal care, expanding healthcare access, and addressing socioeconomic
inequalities—will be essential for mitigating the unintended
consequences of restrictive abortion legislation.

@@

A key limitation of our analysis is the temporal and demographic scope
of the dataset. The data includes only infant deaths under 1 year of age
occurring within the United States to U.S. residents during 2021–2022.
This timeframe does not capture the full year of infant mortality
outcomes following the Dobbs decision, as infants who died in 2023 are
not included. Furthermore, the dataset’s focus on infants under 1 year
of age may exclude cases that reflect longer-term health impacts. These
limitations could lead to an underestimation of the full effects of
abortion bans. Future research should incorporate more comprehensive and
updated datasets to provide a clearer and more complete analysis of the
long-term trends and implications of restrictive abortion policies.

@@

While this study provides valuable findings, several limitations should
be addressed in future research. The analysis relies on aggregate
state-level data, which may obscure localized effects and intra-state
disparities. Unmeasured confounders, such as economic conditions,
healthcare infrastructure differences, or state-specific policies, could
also influence the results. Moreover, the short timeframe of the dataset
limits the ability to assess the long-term effects of abortion policies.
Future research should integrate more granular data, such as
county-level analyses, and extend the study period to capture
longer-term impacts. Qualitative studies focusing on the lived
experiences of affected populations can complement quantitative
analyses, offering a deeper understanding of the broader implications of
restrictive abortion laws. These efforts will support the development of
evidence-based policies aimed at reducing disparities and improving
maternal and infant health outcomes.

@@

@@

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters. In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a crucial role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering insight into the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

@@

In our study, we examine the impact of abortion bans on infant mortality
rates in the United States following the Dobbs decision, employing both
observational data and survey-based research. Observational data provide
quantitative insights, revealing trends and associations across large
populations. However, they often fall short in addressing underlying
mechanisms due to limitations such as confounding factors and challenges
in establishing causality.

>>>>

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "This study examines the impact of the Dobbs decision and subsequent abortion bans on infant mortality rates in the United States, focusing on data from 2021 to 2022. Using a Difference-in-Differences (DID) approach, we analyze trends in states where abortion remained legal compared to those where it became illegal after June 2022. Our results indicate that states enforcing abortion bans experienced an increase of 0.285 infant deaths per 1,000 live births, with compounded effects observed during the post-injunction period. Maternal age and race also emerged as significant predictors, with Black mothers facing disproportionately higher infant mortality rates. These findings underscore the compounded public health implications of restrictive abortion policies and the urgent need for targeted interventions to address racial and demographic disparities. By utilizing recent, high-quality data and rigorous modeling, this study provides essential understanding into the intersection of reproductive rights and infant health outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Instead of examining the impact of abortion laws over a broader time
periods like prior research, our study focuses on the effect of abortion
bans on infant mortality rates in the United States specifically from
2021 to 2022, with an emphasis on changes following the 2022 Dobbs
decision. Using a Difference-in-Differences (DID) approach, we assess
the interaction between abortion legality in states and the timing of
the injunction implemented to understand its influence on infant health
outcomes. The study utilizes observational data from the Centers for
Disease Control and Prevention’s WONDER database, captures infant
mortality trends across states with varying levels of abortion
restrictiveness. Key predictors include the mother’s state of residence,
race, age, and the year of infant death, enabling a detailed examination
of the effects of restrictive abortion laws on public health.

@@

This study is important as we highlight the public health implications
of abortion bans, particularly their role in increasing infant mortality
rates. By focusing on the 2022 Dobbs decision, we provide timely
evidence on how restrictive abortion policies disproportionately affect
vulnerable populations, emphasizing the need for evidence-based policies to
mitigate these adverse outcomes.

@@

The process of translating real-world events into our dataset requires a
systematic approach to measurement and data gathering. In this research,
we analyze infant mortality rates in the United States to understand the
impact of abortion policies and demographic factors. The CDC’s Linked
Birth/Infant Death Records database provides detailed data,
including maternal and infant characteristics, to support this study.
Variables such as maternal age, race, education, birth weight,
gestational age, and prenatal care are included to ensure a detailed
examination of factors influencing infant health outcomes.

@@

Once cleaned, the data is stratified by maternal and infant attributes,
providing understanding into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable understanding, highlighting on
the health implications of varying abortion laws across the U.S. By
utilizing detailed datasets and robust statistical techniques,
this study offers a detailed understanding of the intersection between
policy, demographics, and infant health outcomes, informing the public
health strategies and legislative decisions.

@@

```{r}
#| label: fig-deathrate
#| fig-cap: Distribution of Infant Death Rates: The infant death rate (per 1,000 live births) shows a peak between 5 and 10 deaths per 1,000 infants, with a right-skewed distribution. This skewness highlights that lower infant death rates are more prevalent in the dataset, while higher rates occur less frequently.
#| echo: false

@@

@fig-map illustrates the legality of abortion across U.S. states. Based
on @reproductiveright, after June 2022, abortion has been illegal in
Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
Virginia, and Wisconsin following the June 2022 decision.

@@

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State: After June 2022, abortion is prohibited in the following states: Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-rateandstate
#| fig-cap: Infant Death Rates Over Time: Infant death rates (per 1,000 live births) show a slight increase in states where abortion became illegal following June 2022, whereas rates in states with legal abortion remained relatively stable. This trend suggests a potential association between restrictive abortion laws and higher infant mortality rates.
#| echo: false
#| message: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate \n(per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() +
  theme(legend.position = "bottom")
```

@@

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate by Abortion Legality and Injunction Timing: States categorized as "Illegal Abortion and After Injunction" ('1') exhibit a higher average infant death rate (per 1,000 live births) compared to states in other categories, including "Legal Abortion and Before/After Injunction" or "Illegal Abortion and Before Injunction" ('0'). This highlights a potential link between restrictive abortion laws implemented after the injunction and elevated infant mortality rates.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

@@

The Year of Death variable represents the calendar year in which the
infant death occurred, specifically focusing on the years 2021 and 2022
in this study. This variable is essential for analyzing temporal trends
and identifying patterns or shifts in infant mortality rates over this
essential period. By controlling the year of death, we can closely
examine the impact of significant policy changes, such as the
overturning of Roe v. Wade in 2022, as well as other temporal factors on
infant health outcomes.

@@

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate by Maternal Age Group: Infant mortality rates (per 1,000 live births) vary significantly by maternal age, with younger mothers experiencing higher rates. Mothers aged 20–24 exhibit the highest infant death rate at 10.0, while the lowest rate of 6.6 is observed among mothers aged 30–34, indicating a potential association between maternal age and infant health outcomes.
#| echo: false
#| warning: false
#| message: false
#| 
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

```{r}
#| label: tbl-motherrace
#| tbl-cap: Average Infant Death Rate by Mother’s Single Race: Infant mortality rates (per 1,000 live births) vary significantly by race, with Black or African American mothers experiencing the highest rate at 14.5, compared to 5.8 for White mothers and 5.4 for Asian mothers. This highlights potential disparities in infant health outcomes linked to racial factors.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

```{r}
#| label: fig-choromap
#| fig-cap: Infant Death Rates Across U.S. States (per 1,000 Live Births): The infant mortality rate is significantly higher in the Southeastern states, with Alabama and Mississippi reporting the highest rates. In contrast, states like California and New York exhibit notably lower rates.
#| echo: false

@@

By isolating the impact of the abortion injunction through this
interaction term `after_injunction`×`abortion_illegal`, our model
identifies changes in infant death rates that can be attributed to the
policy intervention. Additional details, validation of model assumption, and diagnostics supporting the
model’s implementation is provided in [Appendix- @sec-model-details].

@@

#### Analysis Framework

For our analysis, we employ Bayesian regression models to estimate infant death rates in states where abortion injunctions were implemented and states where abortion remains legal. The causal effect of the abortion injunction is measured using the interaction term between after_injunction and abortion_illegal. To ensure consistency and comparability, both models share the same predictors, capturing the determinants of infant mortality.

One model includes control variables— `age_of_mother`, `mothers_single_race`, and `year_of_death` to adjust for confounders that may influence infant death rates. This adjustment addresses compositional differences assumed in the DID framework and ensures unbiased estimates of the intervention’s effect. The second model excludes these controls to focus on the overall intervention impact, avoid overfitting, and prioritize simplicity where variability is adequately accounted for by design or random effects. By fitting two models, we compare outcomes to refine estimates and evaluate the direct impact of the intervention.

#### Model Design and Rationale

Our models include state as a random effect to account for regional variations in infant mortality that might introduce bias. Using a random effect rather than a fixed effect captures variability across states without estimating individual effects, modeling unobserved heterogeneity—such as differences in healthcare systems and socioeconomic conditions—through a single variance term. This approach avoids overfitting, enhances efficiency, and ensures generalizability. By adjusting for unmeasured state-specific factors, the models strike a balance between parsimony and interpretability, making them well-suited for our analysis.

#### Why DID Over Alternative Models?

We selected the Difference-in-Differences (DID) framework because randomization is infeasible given the pre-existing differences between states in healthcare systems, socioeconomic conditions, and cultural factors. DID utilizes time and group-level variations, controlling for unobserved, time-invariant confounders. The parallel trends assumption ensures that untreated states provide a valid counterfactual for treated states. Interaction terms, such as after_injunction and abortion_illegal, allow for direct measurement of policy effects, while random effects adjust for regional variability.

Alternative models, such as Fixed Effects, Propensity Score Matching (PSM), and Interrupted Time Series (ITS), were considered but deemed less suitable. Fixed Effects cannot estimate the effects of variables that do not vary over time within a unit. PSM balances observed covariates but does not account for temporal dynamics. ITS analyzes trends within a single group but lacks a comparison group. DID’s ability to integrate temporal and group-level differences makes it the most robust and appropriate framework for estimating the impact of abortion bans on infant mortality.

#### Bayesian Framework

We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically. Using the rstanarm package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale. This improves convergence and ensures priors align with predictor ranges without dominating the data. The Bayesian framework supports plausible parameter centering, captures unobserved heterogeneity through random effects, and ensures robust estimates within the DID framework, assuming parallel trends between groups.

#### Limitations and Applicability

While Bayesian DID models provide robust and interpretable causal estimates, they have limitations. Potential biases from misspecified priors, computational challenges in hierarchical models, and sensitivity to data sparsity or imbalances are notable risks. Additionally, these models may not be suitable if group-level effects are negligible, parallel trends are violated, or data are insufficient. When applied appropriately, however, they offer a reliable approach to understanding the causal effects of abortion bans on infant mortality.

@@

## Model Results

@@

The model highlights key demographic and policy-driven impacts on infant mortality rates. The interaction between abortion bans and the post-injunction period is linked to increased mortality, while states enforcing abortion bans show consistently higher rates. Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions. Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.

@@

The analysis shows distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) shows that abortion bans during the post-injunction
period results in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a essential role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

The findings highlight the significant public health consequences of
abortion bans in the United States. The observed increase of 0.285
infant deaths per 1,000 live births in states enforcing abortion bans
after the Dobbs decision underscores the detrimental impact of
restrictive abortion policies. These increases often reflect systemic
issues such as inadequate prenatal care, gaps in healthcare access, and
underlying socioeconomic inequities. States with pre-existing healthcare
challenges may experience amplified effects, suggesting an urgent need
for detailed policy interventions to address these disparities.
Tackling the root causes of these health disparities—such as improving
prenatal care, expanding healthcare access, and addressing socioeconomic
inequalities—will be essential for mitigating the unintended
consequences of restrictive abortion legislation.

@@

A key limitation of our analysis is the temporal and demographic scope
of the dataset. The data includes only infant deaths under 1 year of age
occurring within the United States to U.S. residents during 2021–2022.
This timeframe does not capture the full year of infant mortality
outcomes following the Dobbs decision, as infants who died in 2023 are
not included. Furthermore, the dataset’s focus on infants under 1 year
of age may exclude cases that reflect longer-term health impacts. These
limitations could lead to an underestimation of the full effects of
abortion bans. Future research should incorporate more detailed and
updated datasets to provide a clearer and more complete analysis of the
long-term trends and implications of restrictive abortion policies.

@@

While this study provides meaningful findings, several limitations should
be addressed in future research. The analysis relies on aggregate
state-level data, which may obscure localized effects and intra-state
disparities. Unmeasured confounders, such as economic conditions,
healthcare infrastructure differences, or state-specific policies, could
also influence the results. Moreover, the short timeframe of the dataset
limits the ability to assess the long-term effects of abortion policies.
Future research should integrate more granular data, such as
county-level analyses, and extend the study period to capture
longer-term impacts. Qualitative studies focusing on the lived
experiences of affected populations can complement quantitative
analyses, offering a deeper understanding of the broader implications of
restrictive abortion laws. These efforts will support the development of
evidence-based policies aimed at reducing disparities and improving
maternal and infant health outcomes.

@@

## Parall Assumption: An Event Study Analysis

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

library(fixest)  # For fast DiD modeling

# Ensure the tidyverse package is loaded
library(tidyverse)

# Define relative_time as numeric (time relative to treatment)
cleaned_data <- analysis_data %>%filter(year_of_death==2022)|>
  mutate(relative_time = (month_code-6))

# Round relative_time to account for fractional months if necessary
cleaned_data <- cleaned_data %>%
  mutate(relative_time = as.factor(round(relative_time, 0)))|>mutate(abortion_illegal=as.numeric(as.character(abortion_illegal)))

est_did <- feols(
  death_rate ~ mothers_single_race + age_of_mother + i(relative_time , abortion_illegal, ref = -1) | state + relative_time, 
  data = cleaned_data
)
```

```{r}
#| echo: false
#| eval: true
#| label: fig-eventstudy
#| tbl-cap: "Event Study Analysis of the Impact of Abortion Bans on Infant Mortality Rates in the United States: Evidence Using June 2022 as the Treatment Date and Pre-Treatment (-1, May) as the Reference Period"
#| warning: false

iplot(est_did,main = "Impact of Abortion Bans on Infant Mortality Rates by State",
  value.lab = "Estimate and __ci__ Conf. Int.",xlab = "Relative Time (Month)")
```

@fig-eventstudy depicts the estimated effects of abortion bans on infant mortality rates over time, with relative months on the x-axis and coefficient estimates with 95% confidence intervals on the y-axis. Before the implementation of abortion bans ($t<0$), the coefficients are centered around zero, indicating no significant differences in infant mortality rates between treated and control states, supporting the parallel trends assumption. Post-ban ($t\geq0$), the estimates show a slight upward trend, suggesting a potential increase in infant mortality rates in restricted states; however, the wide confidence intervals, which overlap with zero, indicate that these effects are not statistically significant. Overall, the results do not provide conclusive evidence of a causal relationship between abortion bans and changes in infant mortality rates, emphasizing the need for further analysis with more precise data or extended time frames.

@@

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters. In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a essential role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering understanding into the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

@@

Our study examines the impact of abortion bans on infant mortality rates in the U.S. following the Dobbs decision, combining observational data and surveys. Observational data reveal trends and associations, but often struggle with confounding factors and causality. Surveys address these gaps by capturing qualitative understanding into individual experiences and contextual factors. This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.
